Likelihoods for a general class of ARGs under the SMC

### 2025-04-23 review-responses.tex

\begin{point}{\revref} % 138:
 is unary in the second tree.
\end{point}

\reply{
Done.
}

\begin{point}{\revref} % 140:
 all lineages with recombination event
\end{point}

\reply{
We  prefer "hit by a recombination event" as it's a little more concrete,
and in keeping with standard phrasing in the field.
}

@@

\begin{point}{\revref{} and a few lines later} % 156/160:
 consolidate how you write backwards in time
\end{point}

\reply{
Done - using "backwards-in-time" now.
}

\begin{point}{\revref{} and elsewhere}
similarly for left-to-right and along-the-genome
\end{point}

\reply{
Done (although we do have ``along the genome'' in places, where
we're not referring to the actual process).
}

@@

\reply{
}

% DISCUSSION

\begin{point}{\revref} % 352:
 modern heuristic ARG ...
\end{point}

\reply{
We prefer to keep this as ``modern ARG inference'' as we feel
that the fact they are heuristic is not really the point---one
could define rigorous methods that also omit full details
about recombination.
}

### 2025-04-23 paper.tex
More review points

\section*{Discussion}
The key contribution of this paper is to rederive a classical
ARG likelihood function in a way that supports the sort of partial
and imprecise information
provided by some modern ARG inference tools, \revpoint{2}{39}
using a novel backwards-time derivation of the SMC.
The underlying process that we reason about probabilistically is still
composed of events affecting lineages, but, because we
do not necessarily observe or estimate these events,
we integrate over the possible timings implied by the information
in the input ARG.
Our method can compute a likelihood under the SMC for a very general class of
ARG~\citep{wong_general_2023},
a significant increase in flexibility over the
strict input requirements for the classical Kuhner-Yamato-Felsenstein  \revpoint{2}{40}
formulation.
Locally unary nodes play a key role, and although not
currently a common focus of ARG research~\citep{wong_general_2023},
they are estimated by tsinfer and may be imputed to a high
degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.
We have shown via some illustrative examples that the formulation is
robust to incomplete and imprecise information, tolerating the loss
of a significant fraction of the internal nodes in the ARG to minimal effect.
We have also shown that the likelihood function performs well
(at least in our toy example) on ARGs inferred by tsinfer+tsdate,
successfully capturing information about latent recombination events
from imprecise and noisy data.
However, there are many more aspects of the likelihood function
that could be examined, beyond coarse parameters like recombination rate
or effective populations size.
% I think we have to grasp this nettle or we're inviting the reviewers to
% give us a bunch of "what if you tried this" bits of extra homework.
Theoretical explanations of this apparent robustness,
and fully characterising the properties of the likelihood on the output
of tsinfer and other methods is an important avenue for future work.
Similarly, while there are many potential applications to concrete
inferential problems, the details are nontrivial and must be
consigned to future work.

@@

Our results, while preliminary, suggest that ``full ARGs'', in which
each recombination and common ancestor event is precisely estimated,
contain a significant degree of redundancy in terms of capturing
information about the generative process.
It is possible, therefore, that explicitly estimating
\emph{all} such events is not necessary to sample ARGs under the
SMC (as ARGweaver and SINGER do).
A combined approach, in which the uncertainty around the
ordering of events that are
weakly informed by the data is encoded structurally in the
graph and other forms of uncertainty captured by sampling from
the posterior, may yield further improvements in
scalability for statistically rigorous inference.
% Signicant work would be needed, however, to make
% downstream analysis robust to such stuctural uncertainty;
% polytomies, for example, are often not
Another exciting prospect for ARG inference is the development of
``hybrid'' inference methods, where heuristic methods such as
tsinfer or ARG-Needle are used to infer
the details of the recent past in huge datasets such as
UK Biobank, passing over to a rigorous model-based approach \revpoint{2}{41}
for the ancient past.
This is similar in spirit to the widely-used approach
of simulating complex dynamics of the recent past using
detailed but slow models and using faster coalescent simulations
in the distant past when the assumptions required
are more justified~\citep{bhaskar_distortion_2014,
kelleher_efficient_2018,haller_tree_2018,nelson_accounting_2020}.
% Maybe we shouldn't bother saying this, but I can't resist the
% opportunity to say "hey, you should be using tskit!"
% It feels like some PIs aren't getting this message, so need to have
% it spelled out.
Just as forward and backward simulations are useful in different contexts
and can be combined to great effect,
large-scale heuristic and Bayesian sampling inference  \revpoint{2}{42}
methods have their applications, and there may be synergy in their combination.
% This is pretty weak, tbh.
The practical details of ``handing over'' between methods
will require cooperation and precise communication;
the tskit library, with its well-defined data model and flexible
metadata support, is the ideal basis for this.

>>>>

\section*{Discussion}
The key contribution of this paper is to rederive a classical
ARG likelihood function in a way that supports the sort of partial
and imprecise information
provided by some modern ARG inference tools, \revpoint{2}{39}
using a novel backwards-time derivation of the SMC.
The underlying process that we reason about probabilistically is still
composed of events affecting lineages, but, because we
do not necessarily observe or estimate these events,
we integrate over the possible timings implied by the information
in the input ARG.
Our method can compute a likelihood under the SMC for a very general class of
ARG~\citep{wong_general_2023},
a significant increase in flexibility over the
strict input requirements for the classical KYF  \revpoint{2}{40}
formulation.
Locally unary nodes play a key role, and although not
currently a common focus of ARG research~\citep{wong_general_2023},
they are estimated by tsinfer and may be imputed to a high
degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.
We have shown via some illustrative examples that the formulation is
robust to incomplete and imprecise information, tolerating the loss
of a significant fraction of the internal nodes in the ARG to minimal effect.
We have also shown that the likelihood function performs well
(at least in our toy example) on ARGs inferred by tsinfer+tsdate,
successfully capturing information about latent recombination events
from imprecise and noisy data.
However, there are many more aspects of the likelihood function
that could be examined, beyond coarse parameters like recombination rate
or effective populations size.
% I think we have to grasp this nettle or we're inviting the reviewers to
% give us a bunch of "what if you tried this" bits of extra homework.
Theoretical explanations of this apparent robustness,
and fully characterising the properties of the likelihood on the output
of tsinfer and other methods is an important avenue for future work.
Similarly, while there are many potential applications to concrete
inferential problems, the details are nontrivial and must be
consigned to future work.

@@

Our results, while preliminary, suggest that ``full ARGs'', in which
each recombination and common ancestor event is precisely estimated,
contain a significant degree of redundancy in terms of capturing
information about the generative process.
It is possible, therefore, that explicitly estimating
\emph{all} such events is not necessary to sample ARGs under the
SMC (as ARGweaver and SINGER do).
A combined approach, in which the uncertainty around the
ordering of events that are
weakly informed by the data is encoded structurally in the
graph and other forms of uncertainty captured by sampling from
the posterior, may yield further improvements in
scalability for statistically rigorous inference.
% Signicant work would be needed, however, to make
% downstream analysis robust to such stuctural uncertainty;
% polytomies, for example, are often not
Another exciting prospect for ARG inference is the development of
``hybrid'' inference methods, where heuristic methods such as
tsinfer or ARG-Needle are used to infer
the details of the recent past in huge datasets such as
UK Biobank, passing over to a rigorous probabilistic approach \revpoint{2}{41}
for the ancient past.
This is similar in spirit to the widely-used approach
of simulating complex dynamics of the recent past using
detailed but slow models and using faster coalescent simulations
in the distant past when the assumptions required
are more justified~\citep{bhaskar_distortion_2014,
kelleher_efficient_2018,haller_tree_2018,nelson_accounting_2020}.
% Maybe we shouldn't bother saying this, but I can't resist the
% opportunity to say "hey, you should be using tskit!"
% It feels like some PIs aren't getting this message, so need to have
% it spelled out.
Just as forward and backward simulations are useful in different contexts
and can be combined to great effect,
large-scale heuristic and probabilistic inference  \revpoint{2}{42}
methods have their applications, and there may be synergy in their combination.
% This is pretty weak, tbh.
The practical details of ``handing over'' between methods
will require cooperation and precise communication;
the tskit library, with its well-defined data model and flexible
metadata support, is the ideal basis for this.
